The effects of context inference on motor learning: savings, de-adaptation and spontaneous recovery

### 2021-09-09 adaptation/article/article.tex
First draft.

\section{Introduction}
It has been long known that humans are capable of learning multiple novel motor tasks at the same time, even when these tasks oppose each other \citep[e.g.][]{Gandolfo_Motor_1996,Shadmehr_Functional_1997}, although interference can occur between tasks \citep[e.g.][]{Brashers-Krug_Consolidation_1996,Sing_Reduction_2010}.

@@

By introducing blocks of trials in which a manipulation is present (e.g. the mechanical arm exerts a force), experimenters are able to observe the dynamics of adaptation through the lens of motor error. A number of phenomena have been observed that persist across participants and across many different motor adaptation experiments: savings, quick de-adaptation, anterograde interference and spontaneous recovery.

In order to explain these phenomena, a number of computational models have been introduced. The least complex of these models are linear learners \citep[e.g.][]{Smith_Interacting_2006,Forano_Timescales_2020,Scheidt_Learning_2001}, in which a single one-dimensional variable is used to describe the adaptation of motor commands given the observed motor errors. By having multiple time scales, each with a learning rate and a forgetting rate, these models have been fitted to experimental data, explaining these distinct experimental phenomena.

@@

In this work, we present a computational model for motor adaptation inspired by models such as MOSAIC \citep{Wolpert_Multiple_1998} and Bayesian switching models \citep{Kording_Bayesian_2004,Oh_Minimizing_2019}. We expand on these models by expanding the concept of context and introducing an explicit mechanism for inferring the context based on Bayesian inference. In our model, the context encompasses all the information relevant to making decisions, both environmental and internal to the agent making these decisions. To infer the current context, our model integrates several sources of information, including the predictions made by its internal forward models, an estimate of the reliability of these predictions, direct sensory information (such as visual contextual cues), proprioceptive information, and feedback provided by the environment. Having inferred the context, the model can choose the best-fitting internal model with which to make decisions.

@@

\subsection{The model}
The model we present has three main components: (1) context-inference, (2) motor adaptation and (3) action selection. Each of these components informs the ones that follow: motor adaptation is informed by context inference, and action selection is informed directly by both motor adaptation and by context inference. A schematic representation of the model can be seen in \fref{fig:model}.

@@

\cite{Oh_Minimizing_2019} posited that the difference between their two groups (~\fref{fig:oh-2019}B and C) can be explained by considering that the smaller visual rotation of 10 degrees caused participants to update their baseline forward model instead of creating a new forward model for the context. If such an account were true, it would predict that when participants walk away from the experiment they would have trouble adapting to the no-rotation reality, having to re-learn it. Our model instead would suggest that participants would immediatly switch back to their baseline movements the moment they walk away from the computer screen, as that is a clear indication that the context has switched back to baseline. More generally, our model predicts that the observed difference between the two groups would disappear if clear sensory contextual cues were added to the experiment. While we believe our model provides a better explanation for the observations by \cite{Oh_Minimizing_2019}, the effect the authors describe has been documented before in chimpanzees \citep{Kojima_Memory_2004}; however, to our knowledge such an effect has not been observed in humans in tasks like that by \cite{Oh_Minimizing_2019}.

The paradigm by \cite{Herzfeld_memory_2014} was designed to test the hypothesis that a memory of errors affected the speed at which participants adapted to curl forces. While it has been shown that sensitivity to error can be modulated \cite{Marko_Sensitivity_2012}, and indeed in our model it depends on the amount of experience (see \sref{sec:interpreting-hyperparameters}, to our knowledge there are no other experiments suggesting that it depends on the history of errors. As before, the introduction of clear sensory contextual cues would allow us to gauge the importance of context inference and of previous errors in the observed data. The same can be said of the experiments by \cite{Davidson_Scaling_2004}, in which different contexts involve movement in different directions, which requires the use of different muscles. This alone could cause delays in switching unrelated to context inference [CITATION NEEDED on the delay in switching muscles?], but we expect the effects to be insufficient to explain the observed data.

>>>>

\section{Introduction}
It has been long known that humans are capable of learning multiple novel motor tasks at the same time, even when these tasks oppose each other \citep[e.g.][]{Gandolfo_Motor_1996,Shadmehr_Functional_1997}, although interference can occur between tasks \citep[e.g.][]{Brashers-Krug_Consolidation_1996,Sing_Reduction_2010}. Additionally, humans have been shown to dynamically switch between different learned tasks. What prompts these switches is still a matter of research.

@@

By introducing blocks of trials in which a manipulation is present (e.g. a mechanical arm exerts a force), experimenters are able to observe the dynamics of adaptation through the lens of motor error. A number of phenomena have been observed that persist across participants and across many different motor adaptation experiments: savings, quick de-adaptation, anterograde interference and spontaneous recovery.

To explain these phenomena, a number of computational models have been introduced. The least complex of these models are linear learners \citep[e.g.][]{Smith_Interacting_2006,Forano_Timescales_2020,Scheidt_Learning_2001}, in which a single one-dimensional variable is used to describe the adaptation of motor commands given the observed motor errors. By having multiple time scales, each with a learning rate and a forgetting rate, these models have been fitted to experimental data, partially explaining these distinct experimental phenomena.

@@

In this work, we show that these unexplained phenomena, alongside savings, de-adaptation, anterograde interference and spontaneous recovery, can be explained in terms of the concept of context, an idea that has recently been gaining traction in different fields \cite[e.g.][]{Sanders_Hippocampal_2020,Hunter_Contextsensitive_2021}.

To this end, we present a computational model for motor adaptation inspired by models such as MOSAIC \citep{Wolpert_Multiple_1998} and Bayesian switching models \citep{Kording_Bayesian_2004,Oh_Minimizing_2019}. We expand on these models by expanding the concept of context and introducing an explicit mechanism for inferring the context based on Bayesian inference. In our model, the context encompasses all the information relevant to making decisions, both environmental and internal to the agent making these decisions. To infer the current context, our model integrates several sources of information, including the predictions made by its internal forward models, an estimate of the reliability of these predictions, direct sensory information (such as visual contextual cues), proprioceptive information, and feedback provided by the environment. Having inferred the context, the model can choose the best-fitting internal model with which to make decisions.

@@

\subsection{Modeling context-dependant adaptation}
The model we present has three main components: (1) context-inference, (2) motor adaptation and (3) action selection. Each of these components informs the ones that follow: motor adaptation is informed by context inference, and action selection is informed directly by both motor adaptation and by context inference. A schematic representation of the model can be seen in \fref{fig:model}.

@@

\cite{Oh_Minimizing_2019} posited that the difference between their two groups (~\fref{fig:oh-2019}B and C) can be explained by considering that the smaller visual rotation of 10 degrees caused participants to update their baseline forward model instead of creating a new forward model for the context. If such an account were true, it would predict that when participants walk away from the experiment they would have trouble adapting to the no-rotation reality, having to re-learn it. Our model instead would suggest that participants would immediatly switch back to their baseline movements the moment they walk away from the computer screen, as that is a clear indication that the context has switched back to baseline. More generally, our model predicts that the observed difference between the two groups would disappear if clear sensory contextual cues were added to the experiment. While we believe our model provides a better explanation for the observations by \cite{Oh_Minimizing_2019}, the effect the authors describe has been documented before in chimpanzees \citep{Kojima_Memory_2004}\todo{@Stefan: you mentioned once you knew of papers talking about this. Do you have any citations I could add?}; however, to our knowledge such an effect has not been observed in humans in tasks like that by \cite{Oh_Minimizing_2019}.

The paradigm by \cite{Herzfeld_memory_2014} was designed to test the hypothesis that a memory of errors affected the speed at which participants adapted to curl forces. While it has been shown that sensitivity to error can be modulated \cite{Marko_Sensitivity_2012}, and indeed in our model it depends on the amount of experience (see \sref{sec:interpreting-hyperparameters}), to our knowledge there are no other experiments suggesting that it depends on the history of errors. As before, the introduction of clear sensory contextual cues would allow us to gauge the importance of context inference and of previous errors in the observed data. The same can be said of the experiments by \cite{Davidson_Scaling_2004}, in which different contexts involve movement in different directions, which requires the use of different muscles. This alone could cause delays in switching unrelated to context inference\todo{Need citation for this}, but we expect the effects to be insufficient to explain the observed data.
